---
title: "Building meaningful machine learning models for disease prediction"
author: "Dr. Shirin Glander"
date: "March 31, 2017"
output:
  html_document: default
  pdf_document:
    keep_tex: yes
---

## Webinar for the [ISDS R Group](http://www.syndromic.org/cop/r)

This document presents the code used to produce the example analysis and figures shown in my webinar on building meaningful machine learning models for disease prediction.

[My webinar slides are available on Github](https://github.com/ShirinG/Webinar\_ML\_for\_disease)

> **Description:** Dr Shirin Glander will go over her work on building machine-learning models to predict the course of different diseases. She will go over building a model, evaluating its performance, and answering or addressing different disease related questions using machine learning. Her talk will cover the theory of machine learning as it is applied using R.

---

### Setup

All analyses are done in R using RStudio. For detailed session information including R version, operating system and package versions, see the `sessionInfo()` output at the end of this document.

All figures are produced with ggplot2.

```{r echo=FALSE, eval=FALSE}
# hypothetical example for describing meaningfulness

library(ggplot2)

df <- data.frame(actual = rep(c("healthy", "disease"), each = 2),
                 predicted = rep(c("healthy", "disease"), 2),
                 count = c(98, 2, 5, 95))

p1 <- ggplot(df, aes(x = predicted, y = count, fill = actual)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Ideal case")

setwd("U:/Github_blog/Webinar/Webinar_ML_for_disease/images")

pdf("images/meaningful_1.pdf", onefile = TRUE, bg = "white", family = "Helvetica", width = 4, height = 3, useDingbats=F)
print(p1)
dev.off()

df <- data.frame(actual = rep(c("healthy", "disease"), each = 2),
                 predicted = rep(c("healthy", "disease"), 2),
                 count = c(60, 40, 5, 95))

p2 <- ggplot(df, aes(x = predicted, y = count, fill = actual)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Possible case 1")

pdf("images/meaningful_2.pdf", onefile = TRUE, bg = "white", family = "Helvetica", width = 4, height = 3, useDingbats=F)
print(p2)
dev.off()

df <- data.frame(actual = rep(c("healthy", "disease"), each = 2),
                 predicted = rep(c("healthy", "disease"), 2),
                 count = c(90, 10, 35, 70))

p3 <- ggplot(df, aes(x = predicted, y = count, fill = actual)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Possible case 2")

pdf("images/meaningful_3.pdf", onefile = TRUE, bg = "white", family = "Helvetica", width = 4, height = 3, useDingbats=F)
print(p3)
dev.off()
```

<br>

### The dataset

The dataset I am using in these example analyses, is the **Breast Cancer Wisconsin (Diagnostic) Dataset**.
The data was downloaded from the [UC Irvine Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29). 

The first dataset looks at the predictor classes:

- malignant or
- benign breast mass.

The features characterise cell nucleus properties and were generated from image analysis of [fine needle aspirates (FNA)](https://en.wikipedia.org/wiki/Fine-needle_aspiration) of breast masses:

- Sample ID (code number)
- Clump thickness
- Uniformity of cell size
- Uniformity of cell shape
- Marginal adhesion
- Single epithelial cell size
- Number of bare nuclei
- Bland chromatin
- Number of normal nuclei
- Mitosis
- Classes, i.e. diagnosis

```{r eval=TRUE}
bc_data <- read.table("datasets/breast-cancer-wisconsin.data.txt", header = FALSE, sep = ",")
colnames(bc_data) <- c("sample_code_number", 
                       "clump_thickness", 
                       "uniformity_of_cell_size", 
                       "uniformity_of_cell_shape", 
                       "marginal_adhesion", 
                       "single_epithelial_cell_size", 
                       "bare_nuclei", 
                       "bland_chromatin", 
                       "normal_nucleoli", 
                       "mitosis", 
                       "classes")

bc_data$classes <- ifelse(bc_data$classes == "2", "benign",
                          ifelse(bc_data$classes == "4", "malignant", NA))
```

<br>

#### Missing data

```{r}
bc_data[bc_data == "?"] <- NA

# how many NAs are in the data
length(which(is.na(bc_data)))

# how many samples would we loose, if we removed them?
nrow(bc_data)
nrow(bc_data[is.na(bc_data), ])
```

<br>

Missing values are imputed with the *mice* package.

```{r eval=FALSE}
# impute missing data
library(mice)

bc_data[,2:10] <- apply(bc_data[, 2:10], 2, function(x) as.numeric(as.character(x)))
dataset_impute <- mice(bc_data[, 2:10],  print = FALSE)
bc_data <- cbind(bc_data[, 11, drop = FALSE], mice::complete(dataset_impute, 1))

bc_data$classes <- as.factor(bc_data$classes)

# how many benign and malignant cases are there?
summary(bc_data$classes)
```

```{r echo=FALSE, eval=FALSE}
save(bc_data, file = "datasets/bc_data.RData")
```

```{r echo=FALSE}
load("datasets/bc_data.RData")
```

<br>

#### Data exploration

- Response variable for classification

```{r response_classification, fig.width=5, fig.height=3}
library(ggplot2)

ggplot(bc_data, aes(x = classes, fill = classes)) +
  geom_bar()
```

<br>

- Response variable for regression

```{r response_regression, fig.width=5, fig.height=3}
ggplot(bc_data, aes(x = clump_thickness)) +
  geom_histogram(bins = 10)
```

<br>

- Principal Component Analysis

```{r pca, fig.width=5, fig.height=3, warning=FALSE, message=FALSE}
library(pcaGoPromoter)
library(ellipse)

# perform pca and extract scores
pcaOutput <- pca(t(bc_data[, -1]), printDropped = FALSE, scale = TRUE, center = TRUE)
pcaOutput2 <- as.data.frame(pcaOutput$scores)
  
# define groups for plotting
pcaOutput2$groups <- bc_data$classes
  
centroids <- aggregate(cbind(PC1, PC2) ~ groups, pcaOutput2, mean)

conf.rgn  <- do.call(rbind, lapply(unique(pcaOutput2$groups), function(t)
  data.frame(groups = as.character(t),
             ellipse(cov(pcaOutput2[pcaOutput2$groups == t, 1:2]),
                   centre = as.matrix(centroids[centroids$groups == t, 2:3]),
                   level = 0.95),
             stringsAsFactors = FALSE)))
    
ggplot(data = pcaOutput2, aes(x = PC1, y = PC2, group = groups, color = groups)) + 
    geom_polygon(data = conf.rgn, aes(fill = groups), alpha = 0.2) +
    geom_point(size = 2, alpha = 0.6) + 
    scale_color_brewer(palette = "Set1") +
    labs(color = "",
         fill = "",
         x = paste0("PC1: ", round(pcaOutput$pov[1], digits = 2) * 100, "% variance"),
         y = paste0("PC2: ", round(pcaOutput$pov[2], digits = 2) * 100, "% variance")) 
```

```{r mds_plot, echo=FALSE, eval=FALSE}
# Multidimensional Scaling 
library(dplyr)
select(bc_data, -1) %>%
  dist() %>%
  cmdscale %>%
  as.data.frame() %>%
  mutate(group = bc_data$classes) %>%
  ggplot(aes(x = V1, y = V2, color = group)) +
    geom_point()
```

```{r tsne_plot, echo=FALSE, eval=FALSE}
# t-SNE dimensionality reduction 
library(tsne)

select(bc_data, -1) %>%
  dist() %>%
  tsne() %>%
  as.data.frame() %>%
  mutate(group = bc_data$classes) %>%
  ggplot(aes(x = V1, y = V2, color = group)) +
    geom_point()
```

- Features

```{r features, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
library(tidyr)

gather(bc_data, x, y, clump_thickness:mitosis) %>%
  ggplot(aes(x = y, color = classes, fill = classes)) +
    geom_density(alpha = 0.3) +
    facet_wrap( ~ x, scales = "free", ncol = 3)
```

```{r warning=FALSE, message=FALSE, fig.width=10, fig.height=5, echo=FALSE, eval=FALSE}
# - Correlation graphs
library(dplyr)
co_mat_benign <- filter(bc_data, classes == "benign") %>%
  select(-1) %>%
  cor()

co_mat_malignant <- filter(bc_data, classes == "malignant") %>%
  select(-1) %>%
  cor()

library(igraph)
g_benign <- graph.adjacency(co_mat_benign,
                         weighted = TRUE,
                         diag = FALSE,
                         mode = "upper")

g_malignant <- graph.adjacency(co_mat_malignant,
                         weighted = TRUE,
                         diag = FALSE,
                         mode = "upper")


# http://kateto.net/networks-r-igraph

cut.off_b <- mean(E(g_benign)$weight)
cut.off_m <- mean(E(g_malignant)$weight)

g_benign_2 <- delete_edges(g_benign, E(g_benign)[weight < cut.off_b])
g_malignant_2 <- delete_edges(g_malignant, E(g_malignant)[weight < cut.off_m])

c_g_benign_2 <- cluster_fast_greedy(g_benign_2) 
c_g_malignant_2 <- cluster_fast_greedy(g_malignant_2) 
```

```{r cor_graph, warning=FALSE, message=FALSE, fig.width=12, fig.height=7, echo=FALSE, eval=FALSE}
par(mfrow = c(1,2))

plot(c_g_benign_2, g_benign_2,
     vertex.size = colSums(co_mat_benign) * 10,
     vertex.frame.color = NA, 
     vertex.label.color = "black", 
     vertex.label.cex = 0.8,
     edge.width	= E(g_benign_2)$weight * 15,
     layout = layout_with_fr(g_benign_2),
     main = "Benign tumors")

plot(c_g_malignant_2, g_malignant_2,
     vertex.size = colSums(co_mat_malignant) * 10,
     vertex.frame.color = NA, 
     vertex.label.color = "black", 
     vertex.label.cex = 0.8,
     edge.width	= E(g_malignant_2)$weight * 15,
     layout = layout_with_fr(g_malignant_2),
     main = "Malignant tumors")
```

<br>

## Machine Learning packages for R

### [caret](http://topepo.github.io/caret/index.html)

```{r warning=FALSE, message=FALSE}
# configure multicore
library(doParallel)
cl <- makeCluster(detectCores())
registerDoParallel(cl)

library(caret)
```

<br>

#### Training, validation and test data

```{r}
set.seed(42)
index <- createDataPartition(bc_data$classes, p = 0.7, list = FALSE)
train_data <- bc_data[index, ]
test_data  <- bc_data[-index, ]
```

```{r distribution, fig.height=4, fig.width=8, fig.align="center", warning=FALSE, message=FALSE}
library(dplyr)

rbind(data.frame(group = "train", train_data),
      data.frame(group = "test", test_data)) %>%
  gather(x, y, clump_thickness:mitosis) %>%
  ggplot(aes(x = y, color = group, fill = group)) +
    geom_density(alpha = 0.3) +
    facet_wrap( ~ x, scales = "free", ncol = 3)
```

<br>

#### Regression

```{r cache=TRUE}
set.seed(42)
model_glm <- caret::train(clump_thickness ~ .,
                          data = train_data,
                          method = "glm",
                          preProcess = c("scale", "center"),
                          trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 10, 
                                                  savePredictions = TRUE, 
                                                  verboseIter = FALSE))
```

```{r echo=FALSE, eval=FALSE}
save(model_glm, file = "models/model_glm.RData")
```

```{r echo=FALSE}
load("models/model_glm.RData")
```

```{r}
model_glm
predictions <- predict(model_glm, test_data)
```

```{r residuals, fig.width=5, fig.height=3}
# model_glm$finalModel$linear.predictors == model_glm$finalModel$fitted.values
data.frame(residuals = resid(model_glm),
           predictors = model_glm$finalModel$linear.predictors) %>%
  ggplot(aes(x = predictors, y = residuals)) +
    geom_jitter() +
    geom_smooth(method = "lm")

# y == train_data$clump_thickness
data.frame(residuals = resid(model_glm),
           y = model_glm$finalModel$y) %>%
  ggplot(aes(x = y, y = residuals)) +
    geom_jitter() +
    geom_smooth(method = "lm")
```

```{r regression_result, fig.width=5, fig.height=3}
data.frame(actual = test_data$clump_thickness,
           predicted = predictions) %>%
  ggplot(aes(x = actual, y = predicted)) +
    geom_jitter() +
    geom_smooth(method = "lm")
```

<br>

#### Classification

##### Decision trees

[rpart](https://cran.r-project.org/web/packages/rpart/rpart.pdf)

```{r decision_tree, cache=TRUE, fig.height=5, fig.width=10, fig.align="center"}
library(rpart)
library(rpart.plot)

set.seed(42)
fit <- rpart(classes ~ .,
            data = train_data,
            method = "class",
            control = rpart.control(xval = 10, 
                                    minbucket = 2, 
                                    cp = 0), 
             parms = list(split = "information"))

rpart.plot(fit, extra = 100)
```

<br>

#### Random Forests

[Random Forests](https://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm) predictions are based on the generation of multiple classification trees. They can be used for both, classification and regression tasks. Here, I show a classification task.

```{r eval=FALSE}
set.seed(42)
model_rf <- caret::train(classes ~ .,
                         data = train_data,
                         method = "rf",
                         preProcess = c("scale", "center"),
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 10, 
                                                  savePredictions = TRUE, 
                                                  verboseIter = FALSE))
```

When you specify `savePredictions = TRUE`, you can access the cross-validation resuls with `model_rf$pred`.

```{r echo=FALSE, eval=FALSE}
save(model_rf, file = "models/model_rf.RData")
```

```{r echo=FALSE}
load("models/model_rf.RData")
```

```{r echo= FALSE, eval=FALSE, fig.width=20, fig.height=10, warning=FALSE, message=FALSE}
str(model_rf$finalModel$forest)

tree <- randomForest::getTree(model_rf$finalModel, 
                              k = 1, 
                              labelVar = TRUE) %>%
  tibble::rownames_to_column() %>%
  mutate(`split point` = ifelse(is.na(prediction), `split point`, NA))

graph_frame <- data.frame(from = rep(tree$rowname, 2),
                          to = c(tree$`left daughter`, tree$`right daughter`))

library(ggraph)
library(igraph)

graph <- graph_from_data_frame(graph_frame) %>%
  delete_vertices("0")

V(graph)$node_label <- gsub("_", " ", as.character(tree$`split var`))
V(graph)$leaf_label <- as.character(tree$prediction)
V(graph)$split <- as.character(round(tree$`split point`, digits = 2))

ggraph(graph, 'dendrogram') + 
  theme_bw() +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = node_label), na.rm = TRUE, repel = TRUE) +
  geom_node_label(aes(label = split), vjust = 2.5, na.rm = TRUE, fill = "white") +
  geom_node_label(aes(label = leaf_label, fill = leaf_label), na.rm = TRUE, repel = TRUE, colour = "white", fontface = "bold", show.legend = FALSE) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank(),
        plot.background = element_rect(fill = "white"),
        panel.border = element_blank(),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size = 18))
  
```

```{r}
model_rf$finalModel$confusion
```

```{r eval=FALSE, echo=FALSE}
model_rf$finalModel$forest$xbestsplit
```

<br>

- Feature Importance

```{r importance_rf, warning=FALSE, message=FALSE, fig.width=5, fig.height=3}
imp <- model_rf$finalModel$importance
imp[order(imp, decreasing = TRUE), ]

# estimate variable importance
importance <- varImp(model_rf, scale = TRUE)
plot(importance)
```

<br>

- predicting test data

```{r}
confusionMatrix(predict(model_rf, test_data), test_data$classes)
```

```{r results_bar_rf, fig.width=5, fig.height=3}
results <- data.frame(actual = test_data$classes,
                      predict(model_rf, test_data, type = "prob"))

results$prediction <- ifelse(results$benign > 0.5, "benign",
                             ifelse(results$malignant > 0.5, "malignant", NA))

results$correct <- ifelse(results$actual == results$prediction, TRUE, FALSE)

ggplot(results, aes(x = prediction, fill = correct)) +
  geom_bar(position = "dodge")
```

```{r results_jitter_rf, fig.width=5, fig.height=3}
ggplot(results, aes(x = prediction, y = benign, color = correct, shape = correct)) +
  geom_jitter(size = 3, alpha = 0.6)
```

<br>

#### Extreme gradient boosting trees

[Extreme gradient boosting (XGBoost)](http://xgboost.readthedocs.io/en/latest/model.html) is a faster and improved implementation of [gradient boosting](https://en.wikipedia.org/wiki/Gradient_boosting) for supervised learning.

> "XGBoost uses a more regularized model formalization to control over-fitting, which gives it better performance." Tianqi Chen, developer of xgboost

XGBoost is a tree ensemble model, which means the sum of predictions from a set of classification and regression trees (CART). In that, XGBoost is similar to Random Forests but it uses a different approach to model training.
Can be used for classification and regression tasks. Here, I show a classification task.

```{r eval=FALSE}
set.seed(42)
model_xgb <- caret::train(classes ~ .,
                          data = train_data,
                          method = "xgbTree",
                          preProcess = c("scale", "center"),
                          trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 10, 
                                                  savePredictions = TRUE, 
                                                  verboseIter = FALSE))
```

```{r echo=FALSE, eval=FALSE}
save(model_xgb, file = "models/model_xgb.RData")
```

```{r echo=FALSE}
load("models/model_xgb.RData")
```

<br>

- Feature Importance

```{r importance_xgb, warning=FALSE, message=FALSE, fig.width=5, fig.height=3}
importance <- varImp(model_xgb, scale = TRUE)
plot(importance)
```

<br>

- predicting test data

```{r}
confusionMatrix(predict(model_xgb, test_data), test_data$classes)
```

```{r results_bar_xgb, fig.width=5, fig.height=3}
results <- data.frame(actual = test_data$classes,
                      predict(model_xgb, test_data, type = "prob"))

results$prediction <- ifelse(results$benign > 0.5, "benign",
                             ifelse(results$malignant > 0.5, "malignant", NA))

results$correct <- ifelse(results$actual == results$prediction, TRUE, FALSE)

ggplot(results, aes(x = prediction, fill = correct)) +
  geom_bar(position = "dodge")
```

```{r results_jitter_xgb, fig.width=5, fig.height=3}
ggplot(results, aes(x = prediction, y = benign, color = correct, shape = correct)) +
  geom_jitter(size = 3, alpha = 0.6)
```

<br>

### Feature Selection

Performing feature selection on the whole dataset would lead to prediction bias, we therefore need to run the whole modeling process on the training data alone!

- Correlation

Correlations between all features are calculated and visualised with the *corrplot* package. I am then removing all features with a correlation higher than 0.7, keeping the feature with the lower mean.

```{r}
library(corrplot)

# calculate correlation matrix
corMatMy <- cor(train_data[, -1])
corrplot(corMatMy, order = "hclust")

#Apply correlation filter at 0.70,
highlyCor <- colnames(train_data[, -1])[findCorrelation(corMatMy, cutoff = 0.7, verbose = TRUE)]

# which variables are flagged for removal?
highlyCor

#then we remove these variables
train_data_cor <- train_data[, which(!colnames(train_data) %in% highlyCor)]
```

<br>

- Recursive Feature Elimination (RFE)

Another way to choose features is with Recursive Feature Elimination. RFE uses a Random Forest algorithm to test combinations of features and rate each with an accuracy score. The combination with the highest score is usually preferential.

```{r echo=FALSE, eval=FALSE}
set.seed(7)
results_rfe <- rfe(x = train_data[, -1], 
                   y = train_data$classes, 
                   sizes = c(1:9), 
                   rfeControl = rfeControl(functions = rfFuncs, method = "cv", number = 10))
```

```{r echo=FALSE, eval=FALSE}
save(results_rfe, file = "models/results_rfe.RData")
```

```{r echo=FALSE}
load("models/results_rfe.RData")
```

```{r }
# chosen features
predictors(results_rfe)

train_data_rfe <- train_data[, c(1, which(colnames(train_data) %in% predictors(results_rfe)))]
```

<br> 

- Genetic Algorithm (GA)

The Genetic Algorithm (GA) has been developed based on evolutionary principles of natural selection: It aims to optimize a population of individuals with a given set of genotypes by modeling selection over time. In each generation (i.e. iteration), each individual's fitness is calculated based on their genotypes. Then, the fittest individuals are chosen to produce the next generation. This subsequent generation of individuals will have genotypes resulting from (re-) combinations of the parental alleles. These new genotypes will again determine each individual's fitness. This selection process is iterated for a specified number of generations and (ideally) leads to fixation of the fittest alleles in the gene pool.

This concept of optimization can be applied to non-evolutionary models as well, like feature selection processes in machine learning.

```{r echo=FALSE, eval=FALSE}
set.seed(27)
model_ga <- gafs(x = train_data[, -1], 
                 y = train_data$classes,
                 iters = 10, # generations of algorithm
                 popSize = 10, # population size for each generation
                 levels = c("malignant", "benign"),
                 gafsControl = gafsControl(functions = rfGA, # Assess fitness with RF
                                           method = "cv",    # 10 fold cross validation
                                           genParallel = TRUE, # Use parallel programming
                                           allowParallel = TRUE))
```

```{r echo=FALSE, eval=FALSE}
save(model_ga, file = "models/model_ga.RData")
```

```{r echo=FALSE}
load("models/model_ga.RData")
```

```{r }
plot(model_ga) # Plot mean fitness (AUC) by generation

train_data_ga <- train_data[, c(1, which(colnames(train_data) %in% model_ga$ga$final))]
```

<br>

### Grid search with caret

- Automatic Grid

```{r eval=FALSE}
set.seed(42)
model_rf_tune_auto <- caret::train(classes ~ .,
                         data = train_data,
                         method = "rf",
                         preProcess = c("scale", "center"),
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 10, 
                                                  savePredictions = TRUE, 
                                                  verboseIter = FALSE,
                                                  search = "random"),
                         tuneLength = 15)
```

```{r echo=FALSE, eval=FALSE}
save(model_rf_tune_auto, file = "models/model_rf_tune_auto.RData")
```

```{r echo=FALSE}
load("models/model_rf_tune_auto.RData")
```

```{r}
model_rf_tune_auto
plot(model_rf_tune_auto)
```

<br>

- Manual Grid

- mtry: Number of variables randomly sampled as candidates at each split.

```{r eval=FALSE}
set.seed(42)
grid <- expand.grid(mtry = c(1:10))

model_rf_tune_man <- caret::train(classes ~ .,
                         data = train_data,
                         method = "rf",
                         preProcess = c("scale", "center"),
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 10, 
                                                  savePredictions = TRUE, 
                                                  verboseIter = FALSE,
                                                  search = "random"),
                         tuneGrid = grid)
```

```{r echo=FALSE, eval=FALSE}
save(model_rf_tune_man, file = "models/model_rf_tune_man.RData")
```

```{r echo=FALSE}
load("models/model_rf_tune_man.RData")
```

```{r}
model_rf_tune_man
plot(model_rf_tune_man)
```

<br>

### Grid search with h2o

The R package h2o provides a convenient interface to [H2O](http://www.h2o.ai/h2o/), which is an open-source machine learning and deep learning platform. H2O distributes a wide range of common machine learning algorithms for classification, regression and deep learning.

```{r warning=FALSE, message=FALSE}
library(h2o)
h2o.init(nthreads = -1)

bc_data_hf <- as.h2o(bc_data)
```

```{r h2o_describe, warning=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.align='center', tidy=FALSE}
h2o.describe(bc_data_hf) %>%
  gather(x, y, Zeros:Sigma) %>%
  mutate(group = ifelse(x %in% c("Min", "Max", "Mean"), "min, mean, max", 
                        ifelse(x %in% c("NegInf", "PosInf"), "Inf", "sigma, zeros"))) %>% 
  ggplot(aes(x = Label, y = as.numeric(y), color = x)) +
    geom_point(size = 4, alpha = 0.6) +
    scale_color_brewer(palette = "Set1") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    facet_grid(group ~ ., scales = "free") +
    labs(x = "Feature",
         y = "Value",
         color = "")
```

```{r corr_plot, warning=FALSE, message=FALSE}
library(reshape2) # for melting

bc_data_hf[, 1] <- h2o.asfactor(bc_data_hf[, 1])

cor <- h2o.cor(bc_data_hf)
rownames(cor) <- colnames(cor)

melt(cor) %>%
  mutate(Var2 = rep(rownames(cor), nrow(cor))) %>%
  mutate(Var2 = factor(Var2, levels = colnames(cor))) %>%
  mutate(variable = factor(variable, levels = colnames(cor))) %>%
  ggplot(aes(x = variable, y = Var2, fill = value)) + 
    geom_tile(width = 0.9, height = 0.9) +
    scale_fill_gradient2(low = "white", high = "red", name = "Cor.") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(x = "", 
         y = "")
```

<br>

#### Training, validation and test data

```{r tidy=FALSE}
splits <- h2o.splitFrame(bc_data_hf, 
                         ratios = c(0.7, 0.15), 
                         seed = 1)

train <- splits[[1]]
valid <- splits[[2]]
test <- splits[[3]]

response <- "classes"
features <- setdiff(colnames(train), response)
```

```{r tidy=FALSE}
summary(train$classes, exact_quantiles = TRUE)
summary(valid$classes, exact_quantiles = TRUE)
summary(test$classes, exact_quantiles = TRUE)
```

```{r pca_features}
pca <- h2o.prcomp(training_frame = train,
           x = features,
           validation_frame = valid,
           transform = "NORMALIZE",
           impute_missing = TRUE,
           k = 3,
           seed = 42)

eigenvec <- as.data.frame(pca@model$eigenvectors)
eigenvec$label <- features

library(ggrepel)
ggplot(eigenvec, aes(x = pc1, y = pc2, label = label)) +
  geom_point(color = "navy", alpha = 0.7) +
  geom_text_repel()
```

<br>

#### Classification

##### Random Forest

```{r eval=FALSE}
hyper_params <- list(
                     ntrees = c(25, 50, 75, 100),
                     max_depth = c(10, 20, 30),
                     min_rows = c(1, 3, 5)
                     )

search_criteria <- list(
                        strategy = "RandomDiscrete", 
                        max_models = 50,
                        max_runtime_secs = 360,
                        stopping_rounds = 5,          
                        stopping_metric = "AUC",      
                        stopping_tolerance = 0.0005,
                        seed = 42
                        )
```

```{r eval=FALSE}
rf_grid <- h2o.grid(algorithm = "randomForest", # h2o.randomForest, 
                                                # alternatively h2o.gbm for Gradient boosting trees
                    x = features,
                    y = response,
                    grid_id = "rf_grid",
                    training_frame = train,
                    validation_frame = valid,
                    nfolds = 25,                           
                    fold_assignment = "Stratified",
                    hyper_params = hyper_params,
                    search_criteria = search_criteria,
                    seed = 42
                    )
```

```{r eval=FALSE}
# performance metrics where smaller is better -> order with decreasing = FALSE
sort_options_1 <- c("mean_per_class_error", "mse", "err", "logloss")

for (sort_by_1 in sort_options_1) {
  
  grid <- h2o.getGrid("rf_grid", sort_by = sort_by_1, decreasing = FALSE)
  
  model_ids <- grid@model_ids
  best_model <- h2o.getModel(model_ids[[1]])
  
  h2o.saveModel(best_model, path="models", force = TRUE)
  
}


# performance metrics where bigger is better -> order with decreasing = TRUE
sort_options_2 <- c("auc", "precision", "accuracy", "recall", "specificity")

for (sort_by_2 in sort_options_2) {
  
  grid <- h2o.getGrid("rf_grid", sort_by = sort_by_2, decreasing = TRUE)
  
  model_ids <- grid@model_ids
  best_model <- h2o.getModel(model_ids[[1]])
  
  h2o.saveModel(best_model, path = "models", force = TRUE)
  
}
```

```{r auc_mse, fig.height=5, fig.width=8}
files <- list.files(path = "models")
rf_models <- files[grep("rf_grid_model", files)]

for (model_id in rf_models) {
  
  #path <- paste0("U:\\Github_blog\\Webinar\\Webinar_ML_for_disease\\models\\", model_id)
  path <- paste0("/Users/Shirin/Documents/Github/Webinar_ML_for_disease/models/", model_id)
  best_model <- h2o.loadModel(path)
  mse_auc_test <- data.frame(model_id = model_id, 
                             mse = h2o.mse(h2o.performance(best_model, test)),
                             auc = h2o.auc(h2o.performance(best_model, test)))
  
  if (model_id == rf_models[[1]]) {
    
    mse_auc_test_comb <- mse_auc_test
    
  } else {
    
    mse_auc_test_comb <- rbind(mse_auc_test_comb, mse_auc_test)
    
  }
}

mse_auc_test_comb %>%
  gather(x, y, mse:auc) %>%
  ggplot(aes(x = model_id, y = y, fill = model_id)) +
    facet_grid(x ~ ., scales = "free") +
    geom_bar(stat = "identity", alpha = 0.8, position = "dodge") +
    scale_fill_brewer(palette = "Set1") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          plot.margin = unit(c(0.5, 0, 0, 1.5), "cm")) +
    labs(x = "", y = "value", fill = "")
```

```{r fig.height=4, fig.width=8, warning=FALSE, message=FALSE}
for (model_id in rf_models) {
  
  best_model <- h2o.getModel(model_id)
  
  finalRf_predictions <- data.frame(model_id = rep(best_model@model_id, 
                                                   nrow(test)),
                                    actual = as.vector(test$classes), 
                                    as.data.frame(h2o.predict(object = best_model, 
                                                              newdata = test)))
  
  finalRf_predictions$accurate <- ifelse(finalRf_predictions$actual == finalRf_predictions$predict, 
                                         "yes", "no")
  
  finalRf_predictions$predict_stringent <- ifelse(finalRf_predictions$benign > 0.8, 
                                                  "benign", 
                                                  ifelse(finalRf_predictions$malignant > 0.8, 
                                                         "malignant", "uncertain"))
  
  finalRf_predictions$accurate_stringent <- ifelse(finalRf_predictions$actual == finalRf_predictions$predict_stringent, 
                                                   "yes", 
                                         ifelse(finalRf_predictions$predict_stringent == "uncertain", 
                                                "na", "no"))
  
  if (model_id == rf_models[[1]]) {
    
    finalRf_predictions_comb <- finalRf_predictions
    
  } else {
    
    finalRf_predictions_comb <- rbind(finalRf_predictions_comb, finalRf_predictions)
    
  }
}
```

```{r final_predictions_rf, fig.height=3, fig.width=9}
finalRf_predictions_comb %>%
  ggplot(aes(x = actual, fill = accurate)) +
    geom_bar(position = "dodge") +
    scale_fill_brewer(palette = "Set1") +
    facet_wrap(~ model_id, ncol = 3) +
    labs(fill = "Were\npredictions\naccurate?",
         title = "Default predictions")

finalRf_predictions_comb %>%
  subset(accurate_stringent != "na") %>%
  ggplot(aes(x = actual, fill = accurate_stringent)) +
    geom_bar(position = "dodge") +
    scale_fill_brewer(palette = "Set1") +
    facet_wrap(~ model_id, ncol = 3) +
    labs(fill = "Were\npredictions\naccurate?",
         title = "Stringent predictions")
```

```{r }
#rf_model <- h2o.loadModel("U:\\Github_blog\\Webinar\\Webinar_ML_for_disease\\models\\rf_grid_model_6")
rf_model <- h2o.loadModel("models/rf_grid_model_6")
#summary(rf_model)
#str(rf_model)
```

```{r }
h2o.varimp_plot(rf_model)
#h2o.varimp(rf_model)
```

```{r }
h2o.mean_per_class_error(rf_model, train = TRUE, valid = TRUE, xval = TRUE)
```

```{r }
h2o.confusionMatrix(rf_model, valid = TRUE)
```

```{r fig.width=8, fig.height=4, fig.align='center'}
plot(rf_model,
     timestep = "number_of_trees",
     metric = "classification_error")
```

```{r fig.width=8, fig.height=4, fig.align='center'}
plot(rf_model,
     timestep = "number_of_trees",
     metric = "logloss")
```

```{r fig.width=8, fig.height=4, fig.align='center'}
plot(rf_model,
     timestep = "number_of_trees",
     metric = "AUC")
```

```{r fig.width=8, fig.height=4, fig.align='center'}
plot(rf_model,
     timestep = "number_of_trees",
     metric = "rmse")
```

```{r }
h2o.auc(rf_model, train = TRUE)
h2o.auc(rf_model, valid = TRUE)
h2o.auc(rf_model, xval = TRUE)
```

```{r fig.width=6, fig.height=5, fig.align='center'}
perf <- h2o.performance(rf_model, test)
perf
```

```{r auc_curve, fig.width=6, fig.height=5, fig.align='center'}
plot(perf)
```

```{r eval=FALSE, echo=FALSE}
perf@metrics$thresholds_and_metric_scores %>%
  ggplot(aes(x = fpr, y = tpr)) +
    geom_point() +
    geom_line() +
    geom_abline(slope = 1, intercept = 0) +
    labs(x = "False Positive Rate",
         y = "True Positive Rate")
```

```{r }
h2o.logloss(perf)
h2o.mse(perf)
h2o.auc(perf)

head(h2o.metric(perf))
```

```{r}
finalRf_predictions <- data.frame(actual = as.vector(test$classes), 
                                  as.data.frame(h2o.predict(object = rf_model, newdata = test)))

finalRf_predictions$accurate <- ifelse(finalRf_predictions$actual == finalRf_predictions$predict, "yes", "no")

finalRf_predictions$predict_stringent <- ifelse(finalRf_predictions$benign > 0.8, "benign", 
                                                ifelse(finalRf_predictions$malignant > 0.8, "malignant", "uncertain"))
finalRf_predictions$accurate_stringent <- ifelse(finalRf_predictions$actual == finalRf_predictions$predict_stringent, "yes", 
                                       ifelse(finalRf_predictions$predict_stringent == "uncertain", "na", "no"))

finalRf_predictions %>%
  group_by(actual, predict) %>%
  dplyr::summarise(n = n())

finalRf_predictions %>%
  group_by(actual, predict_stringent) %>%
  dplyr::summarise(n = n())
```

```{r}
finalRf_predictions %>%
  ggplot(aes(x = actual, fill = accurate)) +
    geom_bar(position = "dodge") +
    scale_fill_brewer(palette = "Set1") +
    labs(fill = "Were\npredictions\naccurate?",
         title = "Default predictions")

finalRf_predictions %>%
  subset(accurate_stringent != "na") %>%
  ggplot(aes(x = actual, fill = accurate_stringent)) +
    geom_bar(position = "dodge") +
    scale_fill_brewer(palette = "Set1") +
    labs(fill = "Were\npredictions\naccurate?",
         title = "Stringent predictions")
```

```{r prop_table, fig.height=3, fig.width=6}
df <- finalRf_predictions[, c(1, 3, 4)]

thresholds <- seq(from = 0, to = 1, by = 0.1)

prop_table <- data.frame(threshold = thresholds, prop_true_b = NA, prop_true_m = NA)

for (threshold in thresholds) {
  pred <- ifelse(df$benign > threshold, "benign", "malignant")
  pred_t <- ifelse(pred == df$actual, TRUE, FALSE)
  
  group <- data.frame(df, "pred" = pred_t) %>%
  group_by(actual, pred) %>%
  dplyr::summarise(n = n())
  
  group_b <- filter(group, actual == "benign")
  
  prop_b <- sum(filter(group_b, pred == TRUE)$n) / sum(group_b$n)
  prop_table[prop_table$threshold == threshold, "prop_true_b"] <- prop_b
  
  group_m <- filter(group, actual == "malignant")
  
  prop_m <- sum(filter(group_m, pred == TRUE)$n) / sum(group_m$n)
  prop_table[prop_table$threshold == threshold, "prop_true_m"] <- prop_m
}

prop_table %>%
  gather(x, y, prop_true_b:prop_true_m) %>%
  ggplot(aes(x = threshold, y = y, color = x)) +
    geom_point() +
    geom_line() +
    scale_color_brewer(palette = "Set1") +
    labs(y = "proportion of true predictions",
         color = "b: benign cases\nm: malignant cases")
```

```{r eval=FALSE}
h2o.shutdown()
```

---

If you are interested in more machine learning posts, check out [the category listing for **machine_learning** on my blog](https://shiring.github.io/categories.html#machine_learning-ref).

------------------

<br>

```{r }
sessionInfo()
```
